From Reversible Programs to \\ Univalent Universes and Back

### 2021-07-09 Pi+/paper/discussion.tex

@@

%%%%%%%%%%%%%%
\paragraph{Free Symmetric Monoidal Category} The forgetful functor from $\SymMonCat$, the 2-category of (small)
symmetric monoidal categories, strong symmetric monoidal functors, and monoidal natural transformations, to the
2-category $\CatCat$, has a left adjoint giving the free symmetric monoidal category $\FSM[\CCat{C}]$ on a category
$\CCat{C}$. This is a 2-monad on $\CatCat$~\cite{blackwellTwodimensionalMonadTheory1989}, whose algebras are (strict)
symmetric monoidal categories. Its construction is known in the literature~\cite{abramskyAbstractScalarsLoops2005}.
Concretely, the objects of $\FSM[\CCat{C}]$ are given by lists of objects of $\CCat{C}$, that is, a pair $(n:\Nat, A:[n]
    \to \CCat{C}_{0})$. An morphism between $(n,A)$ and $(n,B)$ is given by a pair $(\pi,\lambda)$ where $\pi$ is a
permutation of $[n]$, and $\lambda_{i} : A_{i} \to B_{\pi(i)}$ for $1 \leq i \leq n$. Abstractly, this is given by the
Grothendieck construction $\int F$ of the functor $F : \BFin \to \CatCat$ from the groupoid of finite sets and
bijections to $\CatCat$, assigning each natural number $n$ to the $n$-power $C^{n}$ of $C$, and each permutation on
$[n]$ inducing an endofunctor on $C^{n}$ by action. $\BFin$ is the free symmetric monoidal category (groupoid) on one
generator, $\FSM[\unit]$. The free symmetric monoidal category has been used to study
concurrency~\cite{hylandSymmetricMonoidalSketches2004}, petri nets~\cite{baezCategoriesNets2021}, combinatorial
structures~\cite{fioreCartesianClosedBicategory2008}, quantum mechanics~\cite{abramskyAbstractScalarsLoops2005},
bicategorical models of (differential) linear logic~\cite{melliesTemplateGamesDifferential2019}.

Coherence and normalisation problems for monoids in constructive type theory using coherence for monoidal categories was
studied in~\cite{beylinExtractingProofCoherence1996}. In HoTT, coherence for the free monoidal groupoid over a groupoid
and the proof of its universal property has been considered in~\cite{piceghelloCoherenceMonoidalGroupoids2020}.

Free commutative monoids in type theory have been studied in~\cite{gylterudMultisetsTypeTheory2020}, and using HoTT
in~\cite{choudhuryFinitemultisetConstructionHoTT2019}. The free symmetric monoidal groupoid $\FSM[A]$ over a groupoid
$A$ can be given by $\dsum*{X:\UFin}{A^{X}}$, or it can be presented as an algebraic 2-theory using 1-HITs. These HITs
and the proof of their universal property have been considered
in~\cite*{piceghelloCoherenceSymmetricMonoidal2019,choudhuryFinitemultisetConstructionHoTT2019}.

The proof of the universal property of the $\FSM$ is asserted by appealing to Mac Lane's coherence theorem for symmetric
monoidal categories, and using the fact that the finite symmetric group $\Sn$ encodes the permutation group on a finite
set. The existence of the proof is folklore, and we have produced a new proof of it while working in constructive type
theory.

>>>>

%%%%%%%%%%%%%%
\paragraph{Free Symmetric Monoidal Category} The free symmetric monoidal category has been used to study
concurrency~\cite{hylandSymmetricMonoidalSketches2004}, Petri nets~\cite{baezCategoriesNets2021}, combinatorial
structures~\cite{fioreCartesianClosedBicategory2008}, quantum mechanics~\cite{abramskyAbstractScalarsLoops2005}, and
bicategorical models of (differential) linear logic~\cite{melliesTemplateGamesDifferential2019}. The forgetful functor
from $\SymMonCat$, the 2-category of (small) symmetric monoidal categories, strong symmetric monoidal functors, and
monoidal natural transformations, to the 2-category $\CatCat$, has a left adjoint giving the free symmetric monoidal
category $\FSM[\CCat{C}]$ on a category $\CCat{C}$. This is a 2-monad on
$\CatCat$~\cite{blackwellTwodimensionalMonadTheory1989}, whose algebras are (strict) symmetric monoidal categories. Its
construction is known in the literature~\cite{abramskyAbstractScalarsLoops2005}.  Concretely, the objects of
$\FSM[\CCat{C}]$ are given by lists of objects of $\CCat{C}$, that is, a pair $(n:\Nat, A:[n] \to \CCat{C}_{0})$;
morphisms between $(n,A)$ and $(n,B)$ are pairs $(\pi,\lambda)$ where $\pi$ is a permutation of $[n]$, and
$\lambda_{i} : A_{i} \to B_{\pi(i)}$ for $1 \leq i \leq n$. Abstractly, this is given by the Grothendieck construction
$\int F$ of the functor $F : \BFin \to \CatCat$ from the groupoid of finite sets and bijections to $\CatCat$, assigning
each natural number $n$ to the $n$-power $C^{n}$ of $C$, and each permutation on $[n]$ inducing an endofunctor on
$C^{n}$ by action. The groupoid $\BFin$ is the free symmetric monoidal category (groupoid) on one generator, $\FSM[\unit]$.

Coherence and normalisation problems for monoids in constructive type theory using coherence for monoidal categories
were studied by \citet{beylinExtractingProofCoherence1996}. In HoTT, coherence for the free monoidal groupoid over a
groupoid and the proof of its universal property has been considered
by~\citet{piceghelloCoherenceMonoidalGroupoids2020}. Free commutative monoids in type theory have been studied by
\citet{gylterudMultisetsTypeTheory2020}, and using HoTT by \citet{choudhuryFinitemultisetConstructionHoTT2019}. The free
symmetric monoidal groupoid $\FSM[A]$ over a groupoid $A$ can be given by $\dsum*{X:\UFin}{A^{X}}$, or it can be
presented as an algebraic 2-theory using 1-HITs. These HITs and the proof of their universal property have been
considered by \citet{piceghelloCoherenceSymmetricMonoidal2019,choudhuryFinitemultisetConstructionHoTT2019}. The proof of
the universal property of $\FSM$ is asserted by appealing to Mac Lane's coherence theorem for symmetric monoidal
categories, and using the fact that the finite symmetric group $\Sn$ encodes the permutation group on a finite set. The
existence of the proof is folklore. We have produced a proof and formalised it in constructive type theory.

@@
